HO unification from object language to meta language

### 2024-03-29 paper/main.tex
update tex

\begin{invariant}
  Any term in the substitution of \Ho is in normal form.
  TODO: move this to the right place
  TODO: definire normal form, maybe in definition 2.7
  \label{inv:subst-nf}
\end{invariant}

@@

Finally, we can derive the implementation for \maybeeta detection:

@@

Finally, the \maybeeta detection defined just before is an
over-approximation, in the sense that there exists some terms $t$ considered as
\maybeeta, such that forall substitution $\sigma$, $\sigma t$ is not an
$\eta$-expansion. A small example is $\lambda x. f \appsep A_x \appsep A_x$.

The implementation we propose for the \maybeeta relation is given below.

\input{code/maybe_eta}

@@

This rule, to be inserted just before rule~\ref{rule:complam} from the code in
\cref{sec:compilation}, verifies if the \Fo{} term $t$ received in entry is a
\maybeeta. Let $\lambda t'$ be the compiled version of $t$, then the fresh
variable \elpiIn{A} returned as the new \Ho term as in scope all the free names
in $t'$. The critical part of this compilation is the creation of the \linketa,
which links the variable \elpiIn{A} with $t$. This link creation enforce
\cref{inv:linklhs} and \cref{inv:link-eta-right}, since clearly the \lhs is a
variable and the \rhs is a term starting with the \elpiIn{lam} constructor. We
can also note that for any name $x$, $tx = t' \appsep x$ never start with
\elpiIn{lam}, since by contradiction, if 1) $tx$ is an $\eta$-expansion, then
the recursive call to \elpiIn{comp} would have replaced $tx$ with a fresh
variable, which contradict the hypothesis and if 2) $t'$ is not an
$\eta$-expansion, then neither $\lambda t'$ is.

@@

In order to activate a \linketa $l$, we need to implement new rules for the
\elpiIn{progress1} predicate. After passing under all the \elpiIn{abs}
constructors of $l$, there are two cases making a \linketa to progress, 1) \lhs
is instantiated to a rigid term 2) \rhs is no more a \maybeeta or it is a term
which can be reduced to a term with rigid head. 
If \lhs is instantiated to a rigid term $t$, by \cref{inv:subst-nf},
we know that $t$ does not contain any \maybeeta. Let $t'$ the right hand side,
if $t$ is a constant or a function application, then, $t'$, which by
construction has \elpiIn{lam} as head, should be an $\eta$-expansion. We are
therefore allowed to unify $\lambda x. t \appsep x$ (the $\eta$-expanded version
of $t$) with $t'$. Finally, if \lhs starts has \elpiIn{lam} as constructor, then
$t$ is not an $\eta$-expansion and therefore, $t$ can be unified with $t'$.

The second way to activate a \linketa is when the \rhs can $\eta$-reduced to a
term $t$ with rigid head, i.e. $t'$ is not a \maybeeta. This means that we can
unify \lhs with $t'$.

Once a \linketa is activated, it can be removed from \linkStore,
otherwise, the link is kept for a further iteration of \elpiIn{progress}. Note
that this link progression enforce
\cref{inv:subst-nf,inv:linklhs,inv:link-eta-right}: we never commit a term in
the \Ho substitution, since we make unification only when we know that the 
terms are no more \maybeeta, and when \lhs is no more a variable or \rhs is 
no more a \maybeeta, the link is removed from \linkStore.

>>>>

@@

We can now derive the implementation for \maybeeta detection:

@@

As a final remark, the \maybeeta detection defined just before is an
over-approximation, in the sense that there exists some terms $t$ considered as
\maybeeta, such that forall substitution $\sigma$, $\sigma t$ is not an
$\eta$-expansion. A small example is $\lambda x. f \appsep A_x \appsep A_x$
which is considered a \maybeeta. Thi is because, we suppose at the same time
that the second $A_x$ can reduce to $x$, to the first $A_x$ is not reducible to
$A_x$. This is however not considered as a problem, since, adding holes to 
the compiled term does not break \cref{prop:simulation}, since each hole is
connected to a \linketa which is activated exactly when the hole is instantiated
the \linketa, as explain below is activated.

% The implementation we propose for the \maybeeta relation is given below.

% \input{code/maybe_eta}

@@

This rule, to be inserted just before rule~\ref{rule:complam} from the code in
\cref{sec:compilation}, verifies if the \Fo{} term $t$ received in entry is a
\maybeeta. Let $\lambda t'$ be the compiled version of $t$, then the fresh
variable \elpiIn{A} returned as the new \Ho term as in scope all the free names
in $t'$. The critical part of this compilation is the creation of the \linketa,
which links the variable \elpiIn{A} with $t$. This link creation enforce
\cref{inv:linklhs} and \cref{inv:link-eta-right}, since \lhs is a trivially a
variable and the \rhs is a term $t$ starting with the \elpiIn{lam} constructor where
$x$ is free in $t$ otherwise $t$ would not have been detected as a \maybeeta.

\begin{corollary}
  The \rhs of any \linketa has exactly one lambda abstraction.
\end{corollary}

\begin{proof}[Proof sketch]
  By contradiction, suppose that a \linketa $l$ has $t$ as \rhs and $t = \lambda
  x.\lambda y.t'_{xy}$. Two cases are to be analyses: 1) $\lambda y.t'_{xy}$ is
  a $\maybeeta$, then, by construction, \rhs would have been replaced with a the
  $\eta$-expansion of fresh variable $v$, which is a contradiction, since
  $\lambda y.t'_{xy} \neq \lambda x.v_{x}$; 2) $\lambda y.t'_{xy}$ is not an a
  \maybeeta, then neither $t$ is, which is a contradiction since \rhs is always
  a \maybeeta by construction.
\end{proof}

@@

In order to activate a \linketa $l$, we need to extend the \elpiIn{progress1} predicate
by adding new rules. After passing under all the \elpiIn{abs}
constructors of $l$, there are two cases making a \linketa to progress, 1) \lhs
is instantiated to a rigid term 2) \rhs is no more a \maybeeta or it is a term
which can be reduced to a term with rigid head. 
If \lhs is instantiated to a rigid term $t$, by \cref{inv:subst-nf},
we know that $t$ does not contain any \maybeeta. Let $t'$ the right hand side,
if $t$ is a constant or a function application, then, $t'$, which by
construction has \elpiIn{lam} as head, should be an $\eta$-expansion. We are
therefore allowed to unify $\lambda x. t \appsep x$ (the $\eta$-expanded version
of $t$) with $t'$. Finally, if $t$ is a term with \elpiIn{lam} as head, then
it is not an $\eta$-expansion and therefore, $t$ can be unified with $t'$.

The second way to activate a \linketa is when the \rhs is no more a \maybeeta or
\rhs can be $\eta$-reduced to a term $t$ with rigid head. In both cases, \lhs is
unified with $t'$.

Once a \linketa is activated, it can be removed from \linkStore,
otherwise, the link is kept for a further iteration of \elpiIn{progress}. Note
that this link progression enforce
\cref{prop:nf,inv:linklhs,inv:link-eta-right}: we never commit a term in
the \Ho substitution, since we make unification only when we know that the 
terms are no more \maybeeta, and when \lhs is no more a variable or \rhs is 
no more a \maybeeta, the link is removed from \linkStore.

An example
